A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-20 paper.tex
Trip intro commented out stuff

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high-level, ARGs describe the sequence of correlated genealogical
trees along the genome, induced by the different paths of
genetic inheritance caused by recombination.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the \emph{outcome} of such a random process or an inference
of the genetic genealogy of a sample of genome sequences.
However, the distinction between process and structure is not clear cut
and subfields use the term
differently (see Appendix XXX).
%% Sentence deleted: point made already
% ARG inference was for many years synonymous [FIXME: this isn't really true,
% what about parsimony ARGs?]
% with estimating realisations
% of the coalescent with recombination (with, it must be said, limited success),
% and the distinction between process and realisation was not especially important.

%% Repeat of first para
% The recent introduction of two methods, \relate~\citep{speidel2019method}
%and \tsinfer~\citep{kelleher2019inferring}, that estimate structures
%sharing many properties of this classical idea of an ARG, but at previously
%unthinkable scale, has led to a surge of interest in the area.
%
%
%Although both \cite{speidel2019method} and \cite{kelleher2019inferring}
%avoid describing the output of the methods as an ARG,
%and we therefore need a more general definition of an ``ARG'' than the classical
%Griffiths formulation.

@@

% What do we do?
In this paper we attempt to address these problems by
providing a precise definition of an ARG that is general enough to encompass
the output of all popular simulation and inference methods.
% We follow the well established principle that it is advantageous
% to decouple the presence of mutations from the genealogical structure
% itself \citep[contra][chap.~3]{gusfield2014recombinatorics}.
We propose a definition of genetic genealogy in terms of specific genomes
and their intervals of genetic inheritance, and refer to this as
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach,
based on recording evolutionary events, which we call
an ``event ARG'', or ``eARG''.
We show that the gARG encoding has a range of advantages over the
eARG approach, being flexible, efficient and, in particular,
allowing us to systematically encode uncertainty about the
time ordering of recombination events.

% We examineV}
% In particular, we argue that the goal of inferring fully precise
% ARGs---a necessity under the eARG encoding---should be
% % weird phrasing for a criticism?
% critically examined
% %
% as part of the process of fully decoupling our ARG data
% structure from any specific generative model.

% and because it is focused on encoding the \emph{outcome}
% of evolutionary events rather than the details of the events themselves,

% In this paper we [turn into proper round up paragraph later]
% \begin{enumerate}
% \item Discuss the classical Ancestral Recombination Graph formulation,
% and show how it derives from (and is limited by) the coalescent with
% recombination.
% \item In the following sections we
% show that, although the ARG-as-a-data-structure perspective is
% now the most common interpretation, the \emph{form} of that
% data structure is still closely linked to the original
% stochastic process, and that this form is inflexible
% in the types of genetic inheritance that can be described,
% computationally inefficient, and
% forces complete (and potentially spurious) precision about
% the location and timing
% of recombination events. % (and coalescent events?)
% \item Suggest a definition of ARGs that is free of these limitations,
% where the formulation is derived from a (normally diploid) pedigree rather than from
% coalescent approximations, and where the genetic
% material which is transmitted is described by annotations attached to
% edges rather than nodes.
% \item Suggest a classification of ARG nodes that helps us to understand
% the properties of the structures inferred by different methods, and illustrate
% with some examples.

% \item This, coupled with
%     the explicit association of genome coordinates with edges, allows
%     us to represent even the bewilderingly complex patterns of relatedness
%     present in modern datasets.

% \end{enumerate}

% \section*{Genome ARGs}\label{eARG}
% In keeping with its original derivation in terms of a stochastic
% process (see Appendix~\ref{sec-arg-history}), the classical definition of an
% ARG data structure is
% in terms of events.
% Nodes represent
% common ancestor or recombination events that occurred in the
% history of some set of sample genomes, and edges represent ancestral
% lineages~\citep{griffiths1996ancestral}.
% In a common ancestor event the inbound lineages are merged into a
% single ancestral lineage, and in a recombination
% event a lineage is split into two independent
% ancestral lineages.
% % The final vital
% % detail is that we associate a breakpoint with each recombination
% % event.
% % This information is sufficient to uniquely
% % define the local genealogical trees at every position along the genome,
% % which is the basic requirement for a data structure encoding
% % genome-wide genetic ancestry.
% As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
% is usually concretely encoded by associating a recombination breakpoint
% (which is null for common ancestor events) and an ordered list of parents
% with each node. We refer to such a  data structure as an
% ``event ARG'' (eARG).
% To recover the local tree at genome coordinate $x$ (the most
% fundamental operation required for an ARG data structure),
% we traverse the graph rootwards from the leaves.
% At a particular
% node $u$, if it has one parent we are at a common ancestor
% node and we follow that parent. If we are at a
% recombination node, $u$ has two parents: if
% $x$ is less than the breakpoint we follow
% the edge to the first parent, and otherwise follow the edge
% to the second parent.
% The order in which parents are listed at a recombination node is
% therefore significant, telling us
% from which parent the segments of genome to the left and right of the breakpoint
% were inherited.

% In practice,
% several methods explicitly associate information with the outbound edges
% of a recombination event
% to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

% TODO update this when later sections have been drafted.
% As we see in sections XXX and YYY there are substantial limitations
% of expressivity and computational performance when using this eARG
% encoding. More fundamentally, however, the general approach
% of classifying nodes according to a discrete set of events
% and associating genome information with those nodes is intrinsically
% limiting. As we saw in the previous section, the original ARG
% definitions were derived directly from the coalescent with recombination
% stochastic process, and the types of ancestry that can be
% described are therefore limited by the assumptions of that
% model.

% JK: Removing stuff about the limitations of the encoding here
% because it's reasonably straightforward to see an extension in
% which the number of breakpoints is in less than the number of
% parents, and it starts muddying the waters a bit.

% The eARG encoding consists of two types of event
% (common ancestor and recombination) but nature is not so parsimonious
% and there are many other processes that will affect the
% ancestry of samples.
% For example, gene conversion between homologous
% chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
% evolutionary force in many species~\cite[e.g.][]{yang2012great}.
% % Some refs for where people are inferring ancestry in the
% % presence of GC? Isn't this what ClonalFrame etc are all about?
% Similarly, the eARG encoding cannot easily represent multiple
% recombinations on a chromosome, or the transmission of multiple
% chromosomes.
% What we can infer about ancestral histories and
% processes should not be limited by the data structure that we
% use to represent such ancestry. The eARG is simply one way
% to \emph{encode} such history, and, as we have shown,
% one that has significant limitations of expressivity
% derived from its historical origins in the idealised model of
% coalescent with recombination.

% \section*{Node vs edge annotations}
% Another aspect of the classical description of an ARG data structure that
% derives from its historical origins as a stochastic
% process is how we encode information about genetic
% inheritance in the graph. The approach taken by Griffiths to
% doing this is elegant: all we need to do is store
% the location of the breakpoint at each recombination event.
% % Storing a breakpoint for each recombinatin event is a very
% % natural and straightforward thing to do when simulating
% % an  ARG process.
% Given the graph topology and these breakpoints, we can then
% recover the
% local trees at every point in the genome
% by tracing rootwards from the leaves and choosing the
% appropriate outbound path at recombination nodes.
% However, there are several significant disadvantages
% to this approach: [TODO be concrete here later
% and give pointers to the section where we discuss the
% alternative.]

% Fortunately, there is a straightforward change to the
% classical eARG encoding that simplifies
% implementations and greatly increases the flexibility
% of both the types and granularity of inheritance information
% that we can represent.
% The main change that we need to make is to associate
% the details of inheritance with \emph{edges} rather than
% as breakpoints associated with recombination events and
% edge ordering rules.
% This is illustrated in Fig~\ref{fig-arg-data-structure}C
% where this alternative edge annotation formulation is
% shown. By associating the interval $(0, x)$ with edge $(d,e)$
% and the interval $(x, L)$ with $(d, f)$ we make the encoding
% unambiguous: we can list the edges in any order, and it
% will describe the same set of local genealogical trees.
% This approach is also more flexible, because any pattern
% of inheritance can be described by these intervals. For
% example, if a gene conversion event occurred where a segment
% between positions $x$ and $y$ was inherited by
% node $d$ from $e$, we would state this by associating
% the intervals $\{[0, x), [y, L)\}$ with $(d,e)$ and
% $\{[x, y)\}$ with $(d,f)$.
% This change is straightforward, involves no loss of information
% and (as we see in later sections)
% greatly increases both the expressivity and computational
% power of ancestral recombination graph data structures.

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a
``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations.
% A ``genome'' in our usage corresponds to the
% haploid genetic material present in a particular individual or cell.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

>>>>

\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high-level, ARGs describe the sequence of correlated genealogical
trees along the genome, induced by the different paths of
genetic inheritance caused by recombination.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, ARGs have come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the \emph{outcome} of such a random process or an inference
of the genetic genealogy of a sample of genome sequences.
However, the distinction between process and structure is not clear cut
and subfields use the term
differently (see Appendix XXX).

@@

% What do we do?
In this paper we attempt to address these problems by
providing a precise definition of an ARG that is general enough to encompass
the output of all popular simulation and inference methods.
We propose a definition of genetic genealogy in terms of specific genomes
and their intervals of genetic inheritance, and refer to this as
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach,
based on recording evolutionary events, which we call
an ``event ARG'', or ``eARG''.
We show that the gARG encoding has a range of advantages over the
eARG approach, being flexible, efficient and, in particular,
allowing us to systematically encode uncertainty about the
time ordering of recombination events.

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a
``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations.
We are not concerned here with mutational processes or observed sequences,
% FIXME topological isn't the right word...
but consider only processes of inheritance,
following the standard practise in coalescent theory.
% A ``genome'' in our usage corresponds to the
% haploid genetic material present in a particular individual or cell.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
